DSCI 445 - Utilizing Machine Learning for Algae Predictions

### 2020-12-09 paper.Rmd
Added preliminary discussion of methods to both slides and paper.

@@

  Even though algae are natural and part of a healthy freshwater ecosystem, the
algae can grow to an extent that it can be harmful to the ecosystem – creating 
anoxic conditions for example, otherwise known as “dead zones” for the aquatic 
fauna. The HABs can produce toxins that can be harmful to not only the aquatic 
ecosystem, but to humans and other animals, such as dogs. Both algae blooms and 
HABs have a wide range of impacts that can not only be local, but they can also
have interstate impacts. This is especially important to note because Lake Erie,
in 2014, had a HAB that was incredibly impactful to the local economy and 
residents. The HAB left half-a-million people without clean water to drink and 
roughly $65 million in economic losses over a two day span of time.

@@

Methods:
How we got the data
+/- day merge
Tried 3 models
Tried 5 variables
Tried day threshold
Train/test split? (⅓ was held out)
Experimental variables were compared using 5-fold cross validation

Results:

>>>>

```{r setup, echo=FALSE}
library(knitr)
opts_chunk$set(engine.path='venv/bin/python3')
```

# Background

@@

  Even though algae are natural and part of a healthy freshwater ecosystem, the
algae can grow to an extent that it can be harmful to the ecosystem – creating 
anoxic conditions for example, otherwise known as “dead zones” for the aquatic 
fauna. The HABs can produce toxins that can be harmful to not only the aquatic 
ecosystem, but to humans and other animals, such as dogs. Both algae blooms and 
HABs have a wide range of impacts that can not only be local, but they can also
have interstate impacts. This is especially important to note because Lake Erie,
in 2014, had a HAB that was incredibly impactful to the local economy and 
residents. The HAB left half-a-million people without clean water to drink and 
roughly \$65 million in economic losses over a two day span of time.

@@

# Methods

## Methods TODO:
How we got the data

## Variable Selection

## Data Manipulation

Tried 5 variables
+/- day merge
Tried day threshold

The insitu data was not collected as frequently as the satellite imagery data.
To remedy this, an experimental approach was taken in which the FIXME data
was shifted by a certain number of days forward and backward. Intuitively, algae
blooms do not drastically alter the characteristics of a lake between consecutive days.
Interpolating surrounding observations with existing data allowed for many more
of the satellite observations to be used, even though there was no insitu data
accompanying the observations.

However, the aforementioned Interpolating approach certainly fails after a
certain range of days. The optimal shift was determined experimentally: Each
model was fit using data shifted from one to ten days in either direction, and
the cross-validation error was used to assess model accuracy. The specifics of
the experimental structure are described below. The model accuracy provided a
picture of the quality of interpolation provided by each shift. If data was
shifted too many days, the prediction accuracy of the algorithm would
deteriorate. FIXME mention optimal shift and include graph.

## Models

The algae bloom data is spatiotemporal; the blooms are recorded in a certain
geographic location at a certain time. This makes the class of linear models
unsuitable for this problem. Linear models make assumptions about the structure
of the errors in the data generating process that are violated by spatiotemporal
data. Spatiotemporal data has errors which are correlated; as observations
closer in space or time will covary(CITE).

To work around the issue of violated assumptions in classical statistical
models, we opted to use three variants of non-parametric, tree based regression
models. The models used were Random Forest, Gradient Boosting and XGBoost. These
models do not make assumptions about the structure of errors and correlations in
the data, meaning spatiotemporal data is usable.

TODO Describe each model?

## Model Selection

Train/test split? ($\frac{1}{3}$ was held out)
Experimental variables were compared using 5-fold cross validation

# Results

### 2020-12-09 slides.Rmd
Added preliminary discussion of methods to both slides and paper.

---
title: "DSCI 445 - Utilizing Machine Learning for Algae Predictions"
author: "Dawson Eliasen, Boston Lee, Miles Austin"
date: "12/7/2020"
output:
  beamer_presentation: default
  slidy_presentation: default
editor_options:
  chunk_output_type: console
---

@@

```{python, include=FALSE}
from plots import *
```

@@

>>>>

---
title: "DSCI 445 - Utilizing Machine Learning for Algae Predictions"
author: "Dawson Eliasen, Boston Lee, Miles Austin"
date: "12/7/2020"
output:
  beamer_presentation: default
---

@@

```{python, include=FALSE}
from plots import *
=======
```{r setup, include=FALSE}
library(tidyverse)
library(tmap)
library(tmaptools)
library(USAboundaries)
library(sf)
knitr::opts_chunk$set(echo = FALSE)
>>>>>>> Stashed changes
```

@@

## Algorithm Choice

- The algae bloom data is spatiotemporal: We have measurements that are potentially correlated over geographic space, that are taken over time
- This means we cannot use any regression-based methods. Assumptions about uncorrelated errors are violated.
- Solution: Non-parametric algorithms

## Algorithm Choice, Cont.

- We chose two non-parametric algorithms:
  - Random Forest
  - Gradient Boosting

## Gradient Boosted Trees

- Given a differentiable loss function $L(\cdot)$, we want to **minimize** the total loss for points in the test set:

$$\sum_{i = 1}^{n} L(y_i, F(x_i))$$

Where $F(y_i, F(x_i))$ is a measure of how **wrong** $F(x_i)$ is in predicting $y_i$.

- This leads to an optimization problem, which (because of differentiability) can be solved by gradient descent.

## Gradient Boosted Trees, cont.

- We can interpret residuals as the **negative gradient** of the total loss defined before.
- Like AdaBoost, we update the prediction function, but now, we update $F(\cdot)$ based on the negative gradient of our total loss.
- This algorithm can handle different loss functions, with different properties, making it very flexible.
- Here, we use the default implementation in `scikit-learn`, which has a least-squares loss function:

$$L(y_i, F(x_i)) = (y_i = F(x_i))^2$$
